On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-03 paper.tex
new vshear figs
started summary

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Tightly-coupled dust particles
  therefore almost follow the gas flow. This is analogous to the
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  define the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, allowing us to constrain the sharpness of 
dust edges that can persist in protoplanetary disks. We also extend
the vertical shear instability (VSI) in locally isothermal disks to include
dust. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. We briefly discuss potential applications of the 
equivalence between isothermal dusty gas and standard hydrodynamics
to provide thermodynamical interpretations of  other dusty
instabilities in protoplanetary disks.  
\end{abstract}

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  %Tightly-coupled dust particles
  % in protoplanetary disks almost follow the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  identify the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, allowing us to constrain the sharpness of 
dust edges that can persist in protoplanetary disks. We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results are consistent with our definition of
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential applications of the 
equivalence between isothermal dusty gas and standard hydrodynamics
in providing thermodynamical interpretations of other dusty
instabilities in protoplanetary disks.  
\end{abstract}

### 2016-08-03 result.tex
started summary

\subsection{Effect of dust-loading}
We first vary the dust-to-gas ratio $\epsilon_0\in[10^{-3},1]$,
fixing the dust thickness to $\Hd=0.99\Hg$ (so that $\epsilon$ is
roughly constant with height) and perturbation
wavenumber to $k\Hg = 30$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixHd} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy (dotted) in a locally isothermal, dusty disk
    with dust thickness $\Hd=0.99\Hg$ and midplane dust-to-gas ratio
    of $\epsilon_0=10^{-3}$ (black) and $\epsilon_0=1$ (blue). 
    \label{compare_vshear_fixHd}
    }
\end{figure}

Fig. \ref{vsi_dust_loading} show unstable
modes for different values of $\epsilon_0$. The eigenvalue
distributions for $\epsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}, consisting
of the roughly horizontal `body modes' and the nearly-vertical
`surface modes'. The latter is due to the imposed vertical boundaries
\citep{barker15}.

@@

The lowest frequency `fundamental' body mode is not the fastest
growing, but they are energetically dominant because the entire disk
column is perturbed \citep[cf. surface modes which only disturb the
  disk boundaries,][]{umurhan16c}. In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$, where most of the disk mass
reside. Notice also the perturbed dust-to-gas ratio,
$\delta\epsilon$, shifts from being maximized at the disk boundaries
to the midplane.

\begin{figure}
%  \includegraphics[scale=0.54, clip=true, trim=0cm 2.5cm 0cm 0cm]{figures/result2d_dg1d-3.ps}\\
%  \includegraphics[scale=0.54]{figures/result2d_dg1.ps} 
\includegraphics[scale=0.32, clip=true, trim=0.5cm 0cm 3cm 0cm]{figures/result2d_dg1d-3.ps}\includegraphics[scale=0.32, clip=true, trim=1.8cm 0cm 0cm 0cm]{figures/result2d_dg1.ps}
  \caption{Fundamental dusty VSI mode in real space for midplane dust-to-gas
    ratio $\epsilon_0=10^{-3}$ (top) and $\epsilon_0=1$
    (bottom). The color scale shows the perturbation to the
    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
    \label{vsi_dust_loading2d}
    }
\end{figure}

@@

\subsection{Effect of dust thickness}

\subsection{Fixed metalicity} 
Here we vary the dust layer thickness $\Hd$ but hold 
$Z \equiv \epsilon_0 \Htilde = \text{constant,}$ 
so that $\Sigma_\mathrm{d}/\Sigma_\mathrm{g}$ is
fixed. This gives the required midplane dust-to-gas ratio.

%zmax = 2 because we have very thin dust layer

We set $Z=0.03$ and consider two disks with $\Hd=0.1\Hg$ and
$\Hd=0.99\Hg$. Fig. \ref{compare_vshear_fixZ} compares the

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixZ} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy (dotted) in a locally isothermal, dusty disk
    with fixed dust-to-gas mass ratio $Z=0.03$ and dust thickness $\Hd=0.1\Hg$
    (black) and $\Hd=0.99\Hg$ (blue). 
    \label{compare_vshear_fixZ}
    }
\end{figure}

\begin{figure}
  \includegraphics[scale=0.32, clip=true, trim=0.5cm 0cm 3cm 0cm]{figures/result2d_Hd1.ps}\includegraphics[scale=0.32, clip=true, trim=1.8cm 0cm 0cm 0cm]{figures/result2d_Hd0d1.ps} 
  \caption{
%dusty VSI mode in real space for midplane dust-to-gas
%    ratio $\epsilon_0=10^{-3}$ (top) and $\epsilon_0=1$
%    (bottom). The color scale shows the perturbation to the
%    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
%    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
    stuff
    \label{result2d_fixZ}
    }
\end{figure}

>>>>

\subsection{Effect of dust-loading}
We first vary the midplane dust-to-gas ratio 
$\epsilon_0\in[10^{-3},1]$, fixing the dust thickness to 
$\Hd=0.99\Hg$. Then  $\epsilon$ is roughly constant with height. We
set the vertical domain to $\zmax=5\Hg$.

We compare in Fig. \ref{compare_vshear_fixHd} the basic state
vertical shear rate, which is destabilizing, and the vertical buoyancy
frequency, which is stabilizing. For the nearly dust-free disk
$\epsilon_0=10^{-3}$ the vertical shear dominates buoyancy for all
$|z|>0$. However, a heavy dust-load with $\epsilon_0=1$ renders the
buoyancy to dominate over vertical shear in the disk atmosphere
$|z|\gtrsim 2.5$. We thus expect instability all heights for
$\epsilon_0=10^{-3}$, but to be restricted to the midplane for
$\epsilon_0=1$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixHd} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disl with midplane dust-to-gas ratio
    of $\epsilon_0=10^{-3}$ (black) and $\epsilon_0=1$ (blue). 
    The dust layer thickness is fixed to $\Hd=0.99\Hg$ so the 
    dust-to-gas ratio $\epsilon$ is approximately constant with
    height. 
    \label{compare_vshear_fixHd}
    }
\end{figure}

Fig. \ref{vsi_dust_loading} show unstable modes for different values
of $\epsilon_0$ with fixed perturbation wavenumber  $k\Hg = 30$. The
eigenvalue distributions for $\epsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}, consisting
of the roughly horizontal `body modes' and the nearly-vertical
`surface modes'. The latter is due to the imposed vertical boundaries
\citep{barker15}.

@@

The lowest frequency `fundamental' body mode is energetically dominant
because the entire disk column is perturbed \citep[cf. surface modes
  which only disturb the disk boundaries,][]{umurhan16c}. In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$, where most of the disk mass
reside. This is consistent with the above comparison of the basic state vertical
shear and buoyancy. Notice also the perturbed dust-to-gas ratio,
$\delta\epsilon$, shifts from being maximized at the disk boundaries
to the midplane.

\begin{figure}
%  \includegraphics[scale=0.54, clip=true, trim=0cm 2.5cm 0cm 0cm]{figures/result2d_dg1d-3.ps}\\
%  \includegraphics[scale=0.54]{figures/result2d_dg1.ps} 
\includegraphics[scale=0.32, clip=true, trim=0.5cm 0cm 3cm 0cm]{figures/result2d_dg1d-3.ps}\includegraphics[scale=0.32, clip=true, trim=1.8cm 0cm 0cm 0cm]{figures/result2d_dg1.ps}
  \caption{Fundamental dusty VSI mode in real space for midplane dust-to-gas
    ratio $\epsilon_0=10^{-3}$ (left) and $\epsilon_0=1$
    (right). The color scale shows the perturbation to the
    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
    \label{vsi_dust_loading2d}
    }
\end{figure}

@@

\subsection{Effect of dust thickness} 
We now vary the dust layer thickness $\Hd$ but hold 
$Z \equiv \epsilon_0 \Hd/\Hg = \text{constant,}$ 
so that $\Sigma_\mathrm{d}/\Sigma_\mathrm{g}$ is
fixed. This then gives the required midplane dust-to-gas ratio.    
Since we will consider thin dust layers, here we use a smaller
domain with $\zmax=2\Hg$ so that $\epsilon$ does not become
too small.

We set $Z=0.03$ and analyse two disks with $\Hd=0.1\Hg$ and 
$\Hd=0.99\Hg$. Fig. \ref{compare_vshear_fixZ} compares the vertical
vhear rate and buoyancy frequency. For $|z|\gtrsim 0.4\Hg$ the two
disks have the same profile with vertical shear dominating over
buoyancy. We thus expect instabilities away from the disk midplane in 
both cases. For $|z|\lesssim 0.4\Hg$, a thin dust 
layer with $\Hd=0.1\Hg$ boosts the vertical shear rate, but the
associated buoyancy is larger still. Thus we expect the mid-plane of
the disk with $\Hd=0.1\Hg$ to be stable.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixZ} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disk 
    with dust-to-gas mass ratio $Z=0.03$ and dust thickness $\Hd=0.1\Hg$
    (black) and $\Hd=0.99\Hg$ (blue). 
    \label{compare_vshear_fixZ}
    }
\end{figure}

Fig. \ref{result2d_fixZ} compares the fastest-growing VSI body modes  
with $k\Hg=30$ for the two cases above.\citepalias[The thinner domain
  adopted here eliminates surface modes, ][]{lin15}.   
We find very similar mode 
frequencies 
\begin{align*}
  \sigma = \begin{cases}
    \left(0.3053\ii - 0.8142\right)h_\mathrm{g}\OmK & \Hd=0.99\Hg, \\
    \left(0.3178\ii - 1.2237\right)h_\mathrm{g}\OmK & \Hd=0.1\Hg,
  \end{cases}
\end{align*}
since the vertical shear profile is similar throughout most of the
disk. However, meridional motions are surpressed near the midplane of
the $\Hd=0.1\Hg$ disk, as expected from the larger buoyancy frequency
as compared to vertical shear there. This leads to a structure
analogous to PPD dead zones: a quiescent midplane between active
surface layers \citep{gammie96}.

\begin{figure}
  \includegraphics[scale=0.32, clip=true, trim=0.5cm 0cm 3cm 0cm]{figures/result2d_Hd1.ps}\includegraphics[scale=0.32, clip=true, trim=1.8cm 0cm 0cm 0cm]{figures/result2d_Hd0d1.ps} 
  \caption{Fastest-growing dusty VSI mode in real space for midplane
    dust layer thickness $\Hd=0.99\Hg$ (left) and $\Hd=0.1\Hg$
    (right). The dust content is fixed to
    $\Sigma_\mathrm{d}=0.03\Sigma_\mathrm{g}$. 
    The color scale shows the perturbation to the
    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$.
    \label{result2d_fixZ}
    }
\end{figure}
